% ============================================================================
% causalGBKMR -- Software Paper for Environmental Health Perspectives (EHP)
%
% First-author dissertation paper introducing the CausalBKMR R commands
% for time-varying environmental mixtures within the causalmixtures package.
%
% EHP Research Article requirements:
%   - Main text: <= 7,000 words (excluding abstract/refs/tables/captions/supp)
%   - Structured abstract: <= 300 words
%   - No official LaTeX template; use clean article class for Overleaf
% ============================================================================
\documentclass[11pt]{article}

% --- Packages ---------------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{authblk}
\usepackage{lineno}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{setspace}
\usepackage{xurl}
\usepackage{tikz}
\usepackage{verbatim}
\usetikzlibrary{arrows.meta,positioning}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!60!black,
  citecolor=blue!60!black,
  urlcolor=blue!60!black
}

% Numeric reference style for EHP-like citations
\usepackage[numbers,super,sort&compress]{natbib}
\bibliographystyle{unsrtnat}

% R code listing
\lstdefinestyle{Rstyle}{
  language=R,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{red!60!black},
  showstringspaces=false,
  breaklines=true,
  frame=single,
  rulecolor=\color{gray!30}
}
\lstset{style=Rstyle}

\newcommand{\todo}[1]{{\color{red}\textbf{[TODO: #1]}}}
\newcommand{\note}[1]{{\color{purple}\textit{[#1]}}}
\newcommand{\pkg}[1]{\texttt{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\fn}[1]{\texttt{#1()}}

\linenumbers
\onehalfspacing
\emergencystretch=3em

% --- Title and authors ------------------------------------------------------
\title{CausalBKMR: R Commands for Causal Inference\\ with Time-Varying Environmental Mixtures}

\author[1]{Yanran Li}
\author[1]{Arce Domingo-Relloso}
\author[1,2]{Zilan Chai}
\author[3]{Ana Navas-Acien}
\author[4]{Brent A.\ Coull}
\author[1,5,*]{Linda Valeri}

\affil[1]{Department of Biostatistics, Columbia University Mailman School of Public Health, New York, NY, USA}
\affil[2]{Microsoft, Redmond, WA, USA}
\affil[3]{Department of Environmental Health Sciences, Columbia University Mailman School of Public Health, New York, NY, USA}
\affil[4]{Department of Biostatistics, Harvard T.H.\ Chan School of Public Health, Boston, MA, USA}
\affil[5]{Department of Epidemiology, Harvard T.H.\ Chan School of Public Health, Boston, MA, USA}
\affil[*]{Corresponding author: \href{mailto:lv2424@columbia.edu}{lv2424@columbia.edu}}

\date{\today}

\begin{document}
\maketitle

% ============================================================================
% STRUCTURED ABSTRACT (BMC software article: Background / Methods / Results / Conclusions)
% ============================================================================
\begin{abstract}
\noindent\textbf{Background.} Environmental health studies increasingly collect repeated measurements of correlated chemical exposures, biomarkers, and health outcomes. These data raise two linked challenges: exposures occur as mixtures that may act nonlinearly and interactively, and time-varying confounders may be affected by earlier exposure while confounding later exposure--outcome associations. Existing R software addresses either environmental mixtures or longitudinal causal inference, but not an automated g-formula implementation of Bayesian kernel machine regression for time-varying mixtures.

\noindent\textbf{Methods.} \pkg{CausalBKMR} implements g-BKMR, which combines the parametric g-formula with Bayesian kernel machine regression to estimate average causal effects comparing full exposure-mixture trajectories. \pkg{CausalBKMR} is being developed as the time-varying mixture module of the broader \pkg{causalmixtures} package, complementing existing commands for BKMR-based causal mediation analysis and multiple-imputation BKMR. The module accepts longitudinal mixture exposures, time-varying confounders, baseline covariates, and continuous or binary outcomes, fits sequential BKMR models for the confounder and outcome processes, and simulates counterfactual outcome distributions under user-specified exposure trajectories. For large continuous-outcome analyses, the module can call a fastBKMR engine through \pkg{fbkmr}.

\noindent\textbf{Results.} We describe the planned \pkg{causalmixtures} R-command workflow, practical diagnostics, and validation studies for continuous outcomes, binary outcomes, and the fastBKMR engine. \todo{Add final simulation summary after results are available: bias, empirical coverage, posterior interval width, and runtime.}

\noindent\textbf{Conclusions.} \pkg{CausalBKMR} will extend \pkg{causalmixtures} to time-varying environmental mixtures under longitudinal confounding. The module fills a gap between mixture-analysis packages and longitudinal g-formula packages by providing a single command family for data preparation, model fitting, counterfactual simulation, diagnostics, and interpretation.

\vspace{0.5em}
\noindent\textbf{Keywords:} Bayesian kernel machine regression; causal inference; environmental mixtures; g-formula; longitudinal data; R software.
\end{abstract}

% ============================================================================
% BACKGROUND
% ============================================================================
\section*{Background}
\label{sec:background}

Environmental epidemiology studies often measure multiple correlated exposures repeatedly over time. In studies of metals, air pollution, endocrine-disrupting chemicals, or dietary contaminants, the scientific question is rarely whether one exposure at one time point is associated with the outcome. Investigators instead ask how a mixture history affects a later health outcome, whether the effect is nonlinear, and whether the same chemical contributes differently across developmental windows or clinical visits.

Longitudinal mixtures create a causal problem that is not solved by adding all repeated exposures and covariates to a conventional regression model. Time-varying covariates can be affected by earlier exposure and can also predict later exposure and the outcome. Conditioning naively on these covariates may block part of the exposure effect or introduce bias, whereas failing to account for them may leave time-varying confounding uncontrolled. Figure~\ref{fig:dag} summarizes the target data structure: baseline covariates $C_0$ influence the exposure mixture $A_t$, time-varying confounders $L_t$, and outcome $Y$; earlier mixtures affect later confounders; and later confounders affect subsequent mixtures and the final outcome.

Several R packages support related pieces of this problem. The \pkg{bkmr} package implements Bayesian kernel machine regression for flexible analysis of correlated exposure mixtures, including nonlinear exposure-response surfaces, interactions, and posterior inclusion probabilities \citep{bobb_bkmr_2018}. The \pkg{qgcomp} and \pkg{gWQS} packages implement index-based mixture methods that estimate interpretable joint mixture effects under more structured regression assumptions \citep{qgcomp_package,gwqs_package}. The \pkg{causalmixtures} package provides a suite of BKMR-based commands for causal mixture analyses, currently organized around BKMR-CMA for causal mediation analysis and BKMR-MI for analyses with multiple imputation \citep{causalmixtures_package}. For longitudinal causal inference, \pkg{gfoRmula} and related g-formula software estimate effects of time-varying treatment strategies under measured time-varying confounding \citep{mcgrath_gformula_2020,gformula_package}, but these tools do not provide a BKMR mixture-response model as the core exposure-response engine. Thus, existing software covers flexible mixture modeling, BKMR-based mediation and missing-data workflows, or longitudinal g-formula estimation, but does not provide an automated implementation of g-BKMR for time-varying environmental mixtures.

\pkg{CausalBKMR} fills this software gap as the planned third module of \pkg{causalmixtures}. The module implements the g-BKMR estimator proposed by \citet{chai_gbkmr} and provides R commands for data preparation, input auditing, sequential BKMR fitting, Monte Carlo g-computation, posterior summaries, and diagnostic output. During development, these commands are documented on a separate \pkg{CausalBKMR} pkgdown site; in the integrated release, the documentation is intended to appear as a top-level \pkg{CausalBKMR} tab within the \pkg{causalmixtures} website.

% ============================================================================
% METHODS
% ============================================================================
\section*{Methods}
\label{sec:methods}

\subsection*{CausalBKMR estimator}
Consider $n$ individuals followed over $T$ visits. At visit $t$, let $A_t = (A_{t1}, \ldots, A_{tp})$ denote the $p$-component exposure mixture and let $L_t$ denote time-varying covariates that may be affected by earlier exposures and may confound future exposure--outcome relationships. Let $C_0$ denote baseline covariates and let $Y$ denote a final continuous or binary outcome. For two full exposure-mixture trajectories $\bar{\mathbf{a}} = (\mathbf{a}_0,\ldots,\mathbf{a}_{T-1})$ and $\bar{\mathbf{a}}^* = (\mathbf{a}^*_0,\ldots,\mathbf{a}^*_{T-1})$, the target estimand is the average causal effect
\[
  ACE = E\{Y(\bar{\mathbf{a}}^*)\} - E\{Y(\bar{\mathbf{a}})\}.
\]
The default contrast in the \pkg{CausalBKMR} command family compares all mixture components at the 75th percentile with all mixture components at the 25th percentile across all visits, although users can define component-specific and time-specific trajectories.

The estimator uses the longitudinal g-formula to express counterfactual outcome means through the observed-data distribution of the outcome and time-varying confounders. \pkg{CausalBKMR} fits this distribution sequentially. For each post-baseline time-varying confounder, the module fits a BKMR mediator model conditional on the exposure and covariate history. It then fits a BKMR outcome model conditional on the exposure and covariate history. Posterior draws from these models are used to simulate counterfactual confounder histories and outcomes under $\bar{\mathbf{a}}$ and $\bar{\mathbf{a}}^*$, producing posterior samples of $ACE$.

BKMR is used because the mixture-response function can be nonlinear and non-additive. In each BKMR model, the kernel component captures the joint effect of the mixture and relevant covariate history, while baseline covariates enter as linear terms. Variable selection can be incorporated through BKMR posterior inclusion probabilities, allowing the fitted outcome model to identify mixture components and time windows that contribute most strongly to the estimated response surface.

\subsection*{fastBKMR engine}
Standard BKMR can become computationally expensive when the sample size is large or when the longitudinal structure requires many sequential mediator models. For large continuous-outcome analyses, \pkg{CausalBKMR} can use \pkg{fbkmr} through the \code{engine = "fastbkmr"} option. The fastBKMR path fits scalable BKMR approximations on subsets and combines posterior prediction across subsets. In the current implementation, this engine is limited to Gaussian BKMR models; binary outcomes therefore use the standard \pkg{bkmr} engine. \todo{Confirm the exact citation and method description for the installed \pkg{fbkmr} engine before submission.}

% ============================================================================
% SOFTWARE IMPLEMENTATION
% ============================================================================
\section*{Software implementation}
\label{sec:software}

\subsection*{R-command workflow}
\pkg{CausalBKMR} is designed as a small command family within \pkg{causalmixtures}. The recommended workflow is to convert user-supplied matrices into the module's wide longitudinal format with \code{prepare\_gbkmr\_data()}, inspect the detected structure with \code{detect\_variable\_patterns()}, and run the analysis with \code{gbkmr\_run()}. The lower-level function \code{run\_gbkmr\_panel()} is exported for advanced users who already understand the internal data layout, but it is not the recommended entry point for applied analyses.

\begin{verbatim}
library(causalmixtures)

prepared <- prepare_gbkmr_data(
  Y = Y,
  Z = Z,
  X = X,
  time_points = 3,
  mixture_components = 6,
  td_covariates = 2,
  baseline_covariates = 4,
  td_covariate_names = c("waist", "glucose")
)

detect_variable_patterns(prepared, T = 3)

fit <- gbkmr_run(
  data = prepared,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = 3,
  engine = "auto",
  a_probs = c(0.25, 0.75),
  iter = 15000,
  K = 1000
)
\end{verbatim}

The input objects are deliberately simple. \code{Y} is an outcome vector, \code{Z} is an $n \times (pT)$ matrix of mixture exposures ordered by time, and \code{X} is an $n \times (L T + B)$ matrix containing time-varying covariates followed by baseline covariates. The preparation command standardizes the column naming convention, optionally log-transforms mixture variables, and stores metadata describing the number of visits, mixture components, time-varying covariates, and baseline covariates. Until the merge into \pkg{causalmixtures} is complete, the same command family is maintained on the development \pkg{CausalBKMR} documentation site.

\subsection*{Model fitting and output}
\code{gbkmr\_run()} audits the prepared data, chooses a fitting engine, fits the sequential mediator and outcome models, performs Monte Carlo g-computation, and returns an object of class \code{gbkmr\_results}. The output includes the posterior mean and credible interval for the causal contrast, counterfactual means under $\bar{\mathbf{a}}$ and $\bar{\mathbf{a}}^*$, variable-importance summaries, detected data structure, convergence diagnostics, raw model objects, and the original analysis settings.

For binary outcomes, users specify \code{outcome\_type = "binary"}. The standard BKMR engine then fits a probit BKMR outcome model and reports the causal effect on the probability scale as a risk difference. If a user requests \code{engine = "fastbkmr"} for a binary outcome, the command stops with an explicit error because the current fastBKMR path is Gaussian-only.

% ============================================================================
% PRACTICAL CONSIDERATIONS
% ============================================================================
\section*{Practical considerations}
\label{sec:practical}

\subsection*{Diagnostics}
Because \pkg{CausalBKMR} fits several Bayesian models within one longitudinal workflow, diagnostics should be treated as part of the analysis rather than as an optional post-processing step. The command prints an input audit before model fitting, including the detected outcome type, number of exposure components, number of time points, detected time-varying covariates, baseline covariates, selected engine, MCMC settings, and intervention contrast. This audit is intended to catch misspecified column ordering or unintended contrast definitions before a long MCMC run begins.

After fitting, the package prints the estimated causal effect, counterfactual means, model sequence, causal assumptions, and convergence checks. The current diagnostic layer reports effective sample size and Geweke statistics for fitted BKMR models. When diagnostics indicate poor mixing or nonstationarity, users should increase the number of MCMC iterations, use a later post-burn-in selection window, inspect trace plots, and consider simplifying highly unstable exposure or covariate histories.

\subsection*{Identifiability assumptions}
The numerical output of \pkg{CausalBKMR} has a causal interpretation only under standard longitudinal causal assumptions. Consistency requires the observed outcome under the observed exposure history to equal the corresponding counterfactual outcome. Sequential exchangeability requires no unmeasured time-varying confounding after conditioning on the measured exposure and covariate history at each visit. Positivity requires that the exposure trajectories being compared are supported within the observed covariate histories. These assumptions are not guaranteed by the command; they must be justified by study design, measurement quality, subject-matter knowledge, and sensitivity analyses.

\subsection*{Choosing exposure contrasts}
The contrast $\bar{\mathbf{a}}$ versus $\bar{\mathbf{a}}^*$ should represent a scientifically meaningful intervention on the mixture history. The default 25th-versus-75th percentile contrast is useful as a standardized summary of the joint mixture effect, but it may not be appropriate for all applications. When components have different toxicological interpretations, users may specify named vectors through \code{a\_vals} and \code{astar\_vals} to define component-specific and time-specific interventions. Contrasts should remain within the empirical support of the observed exposure distribution unless extrapolation is scientifically justified and clearly disclosed.

% ============================================================================
% RESULTS  (this is the application/case study section that I am writing)
% ============================================================================
\section*{Results}
\label{sec:results}

\subsection*{Simulation study for continuous outcomes}
\todo{Insert the final continuous-outcome simulation design: sample sizes, number of visits, number of mixture components, confounder structure, true contrast, number of simulation replicates, MCMC settings, and comparator methods.}

The continuous-outcome simulation is intended to verify that the standard \pkg{bkmr} engine recovers the target g-BKMR estimand under correctly specified longitudinal data-generating mechanisms. The primary quantities to report are bias in the posterior mean of $ACE$, empirical coverage of 95\% credible intervals, interval width, posterior standard deviation, and convergence diagnostics across fitted mediator and outcome models. \todo{Add numerical results once available.}

\subsection*{Simulation study for binary outcomes}
\todo{Insert the final binary-outcome simulation design and clarify whether binary time-varying confounders are included or whether only the outcome is binary.}

For binary outcomes, \pkg{CausalBKMR} uses the standard BKMR engine with a probit outcome model and reports the causal contrast as a risk difference. The binary-outcome simulation should evaluate whether the probability-scale contrast is recovered under realistic event rates and longitudinal confounding. The primary quantities to report are risk-difference bias, empirical coverage, calibration of posterior uncertainty, and the frequency of convergence warnings. \todo{Add numerical results once available.}

\subsection*{Simulation study with fastBKMR}
\todo{Insert the final fastBKMR simulation or benchmark design: sample size, subset number, cores, wall-clock runtime, memory use, and comparison with standard BKMR when feasible.}

The fastBKMR simulation is intended to evaluate the computational gain of the optional \code{engine = "fastbkmr"} path in large continuous-outcome datasets. The primary quantities to report are wall-clock time, memory use when available, posterior mean agreement with the standard engine in a sample size where both are feasible, and stability across choices of \code{n\_subset} and \code{n\_cores}. \todo{Add numerical results once available.}

% ============================================================================
% DISCUSSION
% ============================================================================
\section*{Discussion}
\label{sec:discussion}

\pkg{CausalBKMR} addresses a practical gap in environmental mixtures research and extends the scope of \pkg{causalmixtures}. The current \pkg{causalmixtures} ecosystem focuses on BKMR-CMA and BKMR-MI; adding \pkg{CausalBKMR} creates a three-module framework for BKMR-based causal mixture analyses: mediation, missing data, and time-varying mixtures. Applied investigators have access to mature software for single-time-point mixture analysis and to general software for longitudinal g-formula analyses, but implementing g-BKMR previously required custom code for data reshaping, sequential BKMR model fitting, counterfactual covariate simulation, posterior summarization, and diagnostics. By packaging these steps into R commands within \pkg{causalmixtures}, \pkg{CausalBKMR} lowers the barrier to applying causal mixture methods in longitudinal environmental health studies.

We recommend that users begin with the default \code{engine = "auto"} workflow and a prespecified 25th-versus-75th percentile mixture contrast for the primary analysis. Before interpreting estimates, users should inspect the input audit, confirm that the detected exposure and confounder ordering matches the study design, verify that the chosen contrast is within empirical support, and review convergence diagnostics. For binary outcomes, users should explicitly set \code{outcome\_type = "binary"} and use the standard \pkg{bkmr} engine. For large continuous-outcome analyses, \code{engine = "auto"} will use fastBKMR when available and appropriate, but users should still assess whether the scalable approximation gives stable summaries under alternative subset settings.

The current implementation has limitations. Time-varying confounders are modeled through Gaussian BKMR mediator models, so binary time-varying confounders should be interpreted cautiously. Baseline covariates enter as linear terms rather than as kernel modifiers. The fastBKMR path is currently limited to continuous-outcome analyses because the installed \pkg{fbkmr} implementation is Gaussian-only. Future work will extend support for non-Gaussian time-varying confounders, scalable binary-outcome analyses, richer posterior visualization, and harmonized documentation with the other \pkg{causalmixtures} modules.

% ============================================================================
% CONCLUSIONS
% ============================================================================
\section*{Conclusions}
\label{sec:conclusions}

\pkg{CausalBKMR} provides an automated R-command workflow for estimating causal effects of time-varying environmental mixtures with g-BKMR. As the planned time-varying-mixture module of \pkg{causalmixtures}, it combines data preparation, sequential BKMR fitting, Monte Carlo g-computation, posterior summaries, diagnostics, and optional fastBKMR computation for large continuous-outcome settings. Once the planned simulation studies are finalized, the software paper will document the operating characteristics of the standard continuous-outcome workflow, the binary-outcome workflow, and the scalable fastBKMR workflow.

% ============================================================================
% DECLARATIONS  (BMC required structure)
% ============================================================================
\section*{Declarations}

\subsection*{Funding}
\todo{Grant numbers from each author's institution.}

\subsection*{Availability of data and materials}
The integrated release will be distributed through \pkg{causalmixtures} at \url{https://arcedr.github.io/causalmixtures/}. Development documentation for the \pkg{CausalBKMR} module is currently available at \url{https://lyric98.github.io/CausalBKMR/}. \todo{Confirm final repository and documentation URLs after the merge.}

\subsection*{Abbreviations}
ACE: average causal effect; BKMR: Bayesian kernel machine regression; CrI: credible interval; DAG: directed acyclic graph; MCMC: Markov chain Monte Carlo; PIP: posterior inclusion probability.

\subsection*{Authors' contributions}
\todo{LV and BC conceived the project. ZC developed the original g-BKMR method. AD-R and YL developed the \pkg{CausalBKMR} command family and its integration plan within \pkg{causalmixtures}. YL conducted the software validation and drafted the software implementation section. ANA contributed to study design and interpretation. All authors read and approved the final manuscript.}

\subsection*{Ethics approval and consent to participate}
Not applicable. This article describes statistical software and simulation studies and does not report new human-subject data analysis.

\subsection*{Consent for publication}
Not applicable.

\subsection*{Competing interests}
The authors declare that they have no competing interests.

\subsection*{Acknowledgements}
\todo{Tribes, study staff, funding bodies.}

% ============================================================================
% REFERENCES
% ============================================================================
\bibliography{references}

% ============================================================================
% FIGURES (placed at end for journal submission per BMC convention)
% ============================================================================
\clearpage
\section*{Figures}

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  scale=0.78,
  transform shape,
  node distance=0.9cm and 1.1cm,
  box/.style={draw, rounded corners, align=center, minimum width=1.35cm, minimum height=0.7cm},
  arr/.style={-{Latex[length=2.5mm]}, thick}
]
\node[box] (C0) {$C_0$\\Baseline\\covariates};
\node[box, right=of C0] (A0) {$A_0$\\Mixture\\visit 0};
\node[box, right=of A0] (L1) {$L_1$\\Time-varying\\confounders};
\node[box, right=of L1] (A1) {$A_1$\\Mixture\\visit 1};
\node[box, right=of A1] (L2) {$L_2$\\Time-varying\\confounders};
\node[box, right=of L2] (A2) {$A_2$\\Mixture\\visit 2};
\node[box, right=of A2] (Y) {$Y$\\Outcome};

\draw[arr] (C0) -- (A0);
\draw[arr] (C0) to[bend left=18] (L1);
\draw[arr] (C0) to[bend left=25] (Y);
\draw[arr] (A0) -- (L1);
\draw[arr] (L1) -- (A1);
\draw[arr] (A1) -- (L2);
\draw[arr] (L2) -- (A2);
\draw[arr] (A2) -- (Y);
\draw[arr] (A0) to[bend right=28] (Y);
\draw[arr] (A1) to[bend right=22] (Y);
\draw[arr] (L1) to[bend left=18] (Y);
\draw[arr] (L2) to[bend left=12] (Y);
\end{tikzpicture}
\caption{Conceptual directed acyclic graph for time-varying environmental mixtures with time-varying confounding. Baseline covariates $C_0$ affect exposure mixtures $A_t$, time-varying confounders $L_t$, and the final outcome $Y$. Earlier mixtures can affect later confounders, which in turn can confound later mixture--outcome relationships.}
\label{fig:dag}
\end{figure}

\end{document}
